---
title: "ECON484 project"
author: "Natalie Li"
date: "2026-06-04"
output: pdf_document
---

# Predicting Employment Outcomes among Young Adults: Education, Well-being, and Machine Learning Evidence from PSID-TAS 2019–2023.

## Main difference between this and the original: added K-fold CV functions for CART, RF, and LASSO. This allows us to compare predictive performance of well-being vs education models more robustly. And, I implemented some modifications to apply them to panel data. Moreover, I used lagged well-being indicators to predict current employment, which is more realistic and avoids simultaneity bias.

```{r setup, message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)
library(glmnet)
library(randomForest)
library(rpart)
library(rpart.plot)
library(lmtest)    # coeftest()
library(sandwich)  # vcovCL() for clustered SEs
```

```{r load}
panel <- read.csv("data/J362431/psid_tas_panel.csv")

panel <- panel %>%
  filter(emp1 != 99) %>%
  mutate(
    employed = ifelse(emp1 == 1, 1, 0),
    year     = factor(year)
  )

cat("Person-waves:", nrow(panel),
    "| Persons:", n_distinct(panel$person_id),
    "| Waves:", paste(levels(panel$year), collapse = "/"), "\n")
table(panel$employed, panel$year)
```

```{r cv_helpers}
# -----------------------------------------------------------------------
# K-fold cross-validation helpers
# Folds are assigned at the PERSON level (grouped CV): all observations
# for a given person go into the same fold, preventing data leakage
# across waves of the same individual.
# -----------------------------------------------------------------------

# Assign each unique person to a fold; return fold index per row.
make_folds <- function(person_ids, k, seed = 123) {
  set.seed(seed)
  persons    <- unique(person_ids)
  person_fold <- setNames(sample(rep_len(seq_len(k), length(persons))),
                          persons)
  unname(person_fold[as.character(person_ids)])
}

# K-fold CV accuracy for CART (rpart).
cv_cart <- function(data, person_ids, k, seed = 123) {
  folds <- make_folds(person_ids, k, seed)
  accs  <- sapply(seq_len(k), function(f) {
    train <- data[folds != f, ]
    test  <- data[folds == f, ]
    m    <- rpart(employed ~ ., data = train, method = "class")
    pred <- as.integer(as.character(predict(m, test, type = "class")))
    mean(pred == test$employed)
  })
  mean(accs)
}

# K-fold CV accuracy for Random Forest.
cv_rf <- function(data, person_ids, k, seed = 123) {
  folds <- make_folds(person_ids, k, seed)
  accs  <- sapply(seq_len(k), function(f) {
    set.seed(seed + f)          # reproducible per fold
    train <- data[folds != f, ]
    test  <- data[folds == f, ]
    m    <- randomForest(as.factor(employed) ~ ., data = train, ntree = 300)
    pred <- as.integer(as.character(predict(m, test)))
    mean(pred == test$employed)
  })
  mean(accs)
}

# K-fold CV accuracy for LASSO.
# Outer loop: k person-grouped folds (same split as CART/RF for fair comparison).
# Inner loop: cv.glmnet on the training fold selects lambda (default 10-fold).
# This avoids the cv.glmnet nfolds >= 3 restriction when k = 2.
cv_lasso <- function(x, y, person_ids, k, seed = 123) {
  folds <- make_folds(person_ids, k, seed)
  accs  <- sapply(seq_len(k), function(f) {
    set.seed(seed + f)
    x_tr <- x[folds != f, , drop = FALSE]; y_tr <- y[folds != f]
    x_te <- x[folds == f, , drop = FALSE]; y_te <- y[folds == f]
    fit  <- cv.glmnet(x_tr, y_tr, family = "binomial", alpha = 1)
    prob <- predict(fit, newx = x_te, s = "lambda.1se", type = "response")
    mean(as.integer(prob > 0.5) == y_te)
  })
  mean(accs)
}

K_VALS <- c(2, 5, 10)
```

---

## Well-being

```{r wellbeing_prep}
wb_vars <- c("wb_happiness_lag", "wb_interest_lag", "wb_satisfied_lag",
             "wb_belong_lag", "wb_society_better_lag", "wb_people_good_lag",
             "wb_manage_daily_lag", "wb_trusting_rel_lag",
             "wb_confident_ideas_lag", "wb_life_direction_lag")

panel <- panel %>%
  mutate(across(all_of(wb_vars), ~ifelse(. %in% c(8, 9), NA, .)))

panel <- panel %>%
  rowwise() %>%
  mutate(wellbeing_score = mean(c_across(all_of(wb_vars)), na.rm = TRUE)) %>%
  ungroup()

summary(panel$wellbeing_score)
```

```{r wellbeing_plot}
ggplot(panel, aes(x = wellbeing_score)) +
  geom_histogram(bins = 20) +
  labs(title = "Distribution of Well-being Scores (Prior Wave)",
       x = "Well-being score (avg B18 items, prior wave, 1-6)", y = "Count")

panel %>%
  group_by(employed) %>%
  summarise(avg_wellbeing = mean(wellbeing_score, na.rm = TRUE))

ggplot(panel, aes(x = factor(employed), y = wellbeing_score)) +
  geom_boxplot() +
  labs(x = "Employed (0 = No, 1 = Yes)", y = "Well-being Score (Prior Wave)",
       title = "Prior-Wave Well-being by Employment Status")
```

```{r wellbeing_logit}
wellbeing_logit <- glm(
  employed ~ wellbeing_score + year,
  data = panel, family = binomial
)
coeftest(wellbeing_logit,
         vcov = vcovCL(wellbeing_logit, cluster = ~person_id))
```
Higher well-being have no association with employment status in the subsequent wave.

```{r wellbeing_tree_data}
# Extract person_ids BEFORE renaming so they align with the data rows.
wb_complete <- panel %>%
  select(person_id, employed, year, all_of(wb_vars)) %>%
  na.omit()

wb_person_ids <- wb_complete$person_id   # kept for grouped CV fold assignment

wb_tree_data <- wb_complete %>%
  select(-person_id)

colnames(wb_tree_data) <- c(
  "employed", "Year",
  "Happy", "InterestInLife", "Satisfied",
  "Community", "SocietyBetter", "PeopleGood",
  "ManageResponsibilities", "TrustingRelationships",
  "Confidence", "LifePurpose"
)
```

```{r wellbeing_tree}
wellbeing_tree <- rpart(employed ~ ., data = wb_tree_data, method = "class")
rpart.plot(wellbeing_tree)
printcp(wellbeing_tree)
```
The CART model failed to identify meaningful splits, suggesting that individual well-being indicators alone do not strongly distinguish employed from non-employed young adults.

```{r wellbeing_rf}
set.seed(123)
wellbeing_rf <- randomForest(
  as.factor(employed) ~ .,
  data = wb_tree_data, ntree = 500, importance = TRUE
)
wellbeing_rf
varImpPlot(wellbeing_rf)
```
Some well-being indicators appear moderately important, but predictive power remains weak.

```{r wellbeing_lasso}
x_wb <- model.matrix(employed ~ ., wb_tree_data)[, -1]
y_wb <- wb_tree_data$employed

set.seed(123)
wellbeing_lasso <- cv.glmnet(x_wb, y_wb, family = "binomial", alpha = 1)
plot(wellbeing_lasso)
coef(wellbeing_lasso, s = "lambda.min")
coef(wellbeing_lasso, s = "lambda.1se")
```
lambda.1se selected no predictors.

```{r wellbeing_cv}
# K-fold CV (grouped by person) for k = 2, 5, 10.
# Folds are split at the person level to avoid leakage across waves.
wb_cv <- data.frame(
  k    = K_VALS,
  CART = sapply(K_VALS, cv_cart,  data = wb_tree_data, person_ids = wb_person_ids),
  RF   = sapply(K_VALS, cv_rf,    data = wb_tree_data, person_ids = wb_person_ids),
  LASSO = sapply(K_VALS, cv_lasso, x = x_wb, y = y_wb, person_ids = wb_person_ids)
)

cat("Well-being model CV accuracy\n")
print(round(wb_cv, 3))

# Visualise
wb_cv_long <- tidyr::pivot_longer(wb_cv, -k,
                                  names_to = "Model", values_to = "Accuracy")
ggplot(wb_cv_long, aes(x = factor(k), y = Accuracy,
                       colour = Model, group = Model)) +
  geom_line() + geom_point(size = 2) +
  scale_y_continuous(limits = c(0.5, 1)) +
  labs(title = "Well-being Models: CV Accuracy by k",
       x = "Number of folds (k)", y = "CV Accuracy") +
  theme_minimal()
```

---

## Education

```{r education_prep}
edu_vars <- c("highest_edu", "grade_level", "attended_college",
              "edu_aspiration", "edu_expectation",
               "mother_edu", "father_edu")

panel <- panel %>%
  mutate(
    highest_edu      = ifelse(highest_edu == 99, NA, highest_edu),
    grade_level      = ifelse(grade_level %in% c(0, 99), NA, grade_level),
    attended_college = ifelse(attended_college == 9, NA, attended_college),
    edu_aspiration   = ifelse(edu_aspiration %in% c(8, 98, 99), NA,
                              edu_aspiration),
    edu_expectation  = ifelse(edu_expectation %in% c(0, 9), NA,
                              edu_expectation),
    mother_edu       = ifelse(mother_edu %in% c(0, 96, 99), NA, mother_edu),
    father_edu       = ifelse(father_edu %in% c(0, 96, 98, 99), NA, father_edu)
  )
```

```{r education_logit}
edu_logit_data <- panel %>%
  select(person_id, employed, year, all_of(edu_vars)) %>%
  na.omit()

cat("Education logit sample:", nrow(edu_logit_data), "person-waves\n")

education_logit <- glm(
  employed ~ . - person_id,
  data = edu_logit_data, family = binomial
)
coeftest(education_logit,
         vcov = vcovCL(education_logit, cluster = ~person_id))
```

```{r education_tree_data}
# Extract person_ids BEFORE renaming.
edu_complete <- panel %>%
  select(person_id, employed, year, all_of(edu_vars)) %>%
  na.omit()

edu_person_ids <- edu_complete$person_id

edu_tree_data <- edu_complete %>%
  select(-person_id)

colnames(edu_tree_data) <- c(
  "employed", "Year",
  "HighestEducation", "GradeCompleted",
  "AttendedCollege",
  "EducationAspiration", "EducationExpectation",
  "MotherEducation", "FatherEducation"
)

cat("Education ML sample:", nrow(edu_tree_data), "person-waves\n")
```

```{r education_tree}
education_tree <- rpart(employed ~ ., data = edu_tree_data, method = "class")
rpart.plot(education_tree)
```

```{r education_rf}
set.seed(123)
education_rf <- randomForest(
  as.factor(employed) ~ .,
  data = edu_tree_data, ntree = 500, importance = TRUE
)
varImpPlot(education_rf)
```

```{r education_lasso}
x_edu <- model.matrix(employed ~ ., edu_tree_data)[, -1]
y_edu <- edu_tree_data$employed

set.seed(123)
education_lasso <- cv.glmnet(x_edu, y_edu, family = "binomial", alpha = 1)
plot(education_lasso)
coef(education_lasso, s = "lambda.min")
coef(education_lasso, s = "lambda.1se")
```

```{r education_cv}
edu_cv <- data.frame(
  k     = K_VALS,
  CART  = sapply(K_VALS, cv_cart,  data = edu_tree_data, person_ids = edu_person_ids),
  RF    = sapply(K_VALS, cv_rf,    data = edu_tree_data, person_ids = edu_person_ids),
  LASSO = sapply(K_VALS, cv_lasso, x = x_edu, y = y_edu, person_ids = edu_person_ids)
)

cat("Education model CV accuracy\n")
print(round(edu_cv, 3))

edu_cv_long <- tidyr::pivot_longer(edu_cv, -k,
                                   names_to = "Model", values_to = "Accuracy")
ggplot(edu_cv_long, aes(x = factor(k), y = Accuracy,
                        colour = Model, group = Model)) +
  geom_line() + geom_point(size = 2) +
  scale_y_continuous(limits = c(0.5, 1)) +
  labs(title = "Education Models: CV Accuracy by k",
       x = "Number of folds (k)", y = "CV Accuracy") +
  theme_minimal()
```

---

## Comparison: Well-being vs Education

```{r comparison}
wellbeing_AIC <- AIC(wellbeing_logit)
education_AIC <- AIC(education_logit)

wb_tree_splits  <- wellbeing_tree$cptable[nrow(wellbeing_tree$cptable), "nsplit"]
edu_tree_splits <- education_tree$cptable[nrow(education_tree$cptable), "nsplit"]

wb_lasso_vars  <- sum(coef(wellbeing_lasso,  s = "lambda.1se")[-1, ] != 0)
edu_lasso_vars <- sum(coef(education_lasso, s = "lambda.1se")[-1, ] != 0)

wb_rf_acc  <- mean(predict(wellbeing_rf,  wb_tree_data)  == wb_tree_data$employed)
edu_rf_acc <- mean(predict(education_rf, edu_tree_data) == edu_tree_data$employed)

comparison_table <- data.frame(
  Model_Group          = c("Well-being", "Education"),
  Logit_AIC            = c(wellbeing_AIC, education_AIC),
  Tree_Splits          = c(wb_tree_splits, edu_tree_splits),
  LASSO_1se_Vars       = c(wb_lasso_vars, edu_lasso_vars),
  RF_Training_Accuracy = c(wb_rf_acc, edu_rf_acc)
)

comparison_table
```

```{r cv_comparison}
# Side-by-side CV accuracy table: well-being vs education, for k = 2, 5, 10.
cv_summary <- data.frame(
  k             = K_VALS,
  WB_CART       = wb_cv$CART,
  WB_RF         = wb_cv$RF,
  WB_LASSO      = wb_cv$LASSO,
  Edu_CART      = edu_cv$CART,
  Edu_RF        = edu_cv$RF,
  Edu_LASSO     = edu_cv$LASSO
)

cat("CV accuracy: well-being vs education models\n")
print(round(cv_summary, 3))
```

```{r comparison_plots}
ggplot(comparison_table, aes(x = Model_Group, y = Logit_AIC)) +
  geom_col() +
  labs(title = "Logit Model Fit Comparison", x = "", y = "AIC")

ggplot(comparison_table, aes(x = Model_Group, y = LASSO_1se_Vars)) +
  geom_col() +
  labs(title = "LASSO Variable Selection Comparison",
       x = "", y = "Variables Selected (lambda.1se)")
```
---

## Combined Model: Does Well-being Add Anything to Education?

```{r combined}
combined_data <- panel %>%
  select(person_id, employed, year, wellbeing_score, all_of(edu_vars)) %>%
  na.omit()

colnames(combined_data) <- c(
  "person_id", "employed", "Year", "WellbeingScore",
  "HighestEducation", "GradeCompleted", "AttendedCollege",
  "EducationAspiration", "EducationExpectation",
  "MotherEducation", "FatherEducation"
)

cat("Combined-model sample size:", nrow(combined_data), "person-waves\n")
```

```{r combined_models}
combined_edu_only <- glm(
  employed ~ . - WellbeingScore - person_id,
  data = combined_data, family = binomial
)

combined_full <- glm(
  employed ~ . - person_id,
  data = combined_data, family = binomial
)

coeftest(combined_full,
         vcov = vcovCL(combined_full, cluster = ~person_id))
```

```{r combined_test}
anova(combined_edu_only, combined_full, test = "LRT")
AIC(combined_edu_only, combined_full)
```
